\vspace{-2.5em} Spatio-angular transfer functions for fluorescence microscopes\vspace{-1em}

### 2018-04-11 notes/2018-02-23-spatio-angular-kernel/report/report.tex
Converging on final OTF w/o polarizers. Patrick's ISBI edits.

%%%%%%%%%%%%
% Document %
%%%%%%%%%%%%
\begin{document}
\title{\vspace{-2.5em} Spatio-angular transfer functions for fluorescence microscopes\vspace{-1em}}  \author{Talon Chandler, Min Guo, Hari Shroff, Rudolf Oldenbourg, Patrick La Rivi\`ere}
\date{\vspace{-1em}\today\vspace{-1em}}
\maketitle
\begin{abstract}
  We investigate how the orientation and position of fluorescent dipole emitters
  affects microscopic imaging using electromagnetic optics theory. Starting with
  the thoroughly studied spatio-angular point spread function, we introduce the
  spatio-angular coherent spread function, coherent transfer function, and
  optical transfer function as electromagnetic extensions of well-known
  functions in scalar optics theory. We use these concepts to show that
  fluorescence microscopes have a spatio-angular band limit. Finally, we study
  polarized light microscopes and find that polarized illumination is a type of
  structured illumination that extends the angular band limit.
\end{abstract}
\section{Introduction}
TODO

@@

Figure \ref{fig:frames_a} shows a schematic of the fluorescence microscope that
we are considering with a summary of our notation. We start by following Backer
and Moerner \cite{backer2014} to find the electric field at position $\rb{}$ in
the back focal plane due to a single dipole emitter at position $\ro{}$ oriented
along $\so{}$ as
\begin{align}
  \mb{\tilde{e}}_b(\rb{};\ro{}, \so{}) \propto \me^{-i(kn_o/f_o)\rb{}\cdot\ro{}}\sqrt{\frac{1}{\rho_b}}
  \begin{bmatrix}
    \sin^2\phi_b + \rho_b\cos^2\phi_b&\sin\phi_b\cos\phi_b(\rho_b - 1)&-\frac{r_b}{f_o}\cos\phi_b\\
    \sin\phi_b\cos\phi_b(\rho_b - 1)&\cos^2\phi_b + \rho_b\sin^2\phi_b&-\frac{r_b}{f_o}\sin\phi_b\\
    0&0&0
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix}
\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)\label{eq:bfp}
\end{align}
where we define $\rho_b \equiv \sqrt{1 - \left(\frac{r_b}{f_o}\right)^2}$, and
$\Pi(x)$ is a boxcar function that returns 1 when $|x| < 1$ and 0 otherwise. We
can understand this expression term by term---the exponential term accounts for
the phase as dipole emitter is moved in object space, the square root term
conserves power before and after the objective lens, the matrix models the
dipole emission process and electric field rotation caused by the objective
lens, the vector is the dipole orientation unit vector, and
$\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)$ accounts for the numerical
aperture of the lens with $r_b^{\text{max}} = \frac{f_0}{n_0}\text{NA}$. We use
a tilde to mark single dipole response functions---we will consider the response
due to a field of dipoles in the next section.

Although we will not use the result immediately, we can write the electric field
in back focal plane under the paraxial approximation. We expand Eq. \ref{eq:bfp}
in a Taylor series about $r_b=0$ and drop all second-order and higher terms to
find
\begin{align}
  \mb{\tilde{e}}^{(p)}_b(\rb{};\ro{}, \so{}) \propto \me^{-i(kn_o/f_o)\rb{}\cdot\ro{}}
  \begin{bmatrix}
    1 & 0 &-\frac{r_b}{f_o}\cos\phi_b\\
    0&1&-\frac{r_b}{f_o}\sin\phi_b\\
    0&0&0
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix}
\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)\label{eq:bfppara}
\end{align}
We will continue to use a superscript $(p)$ to mark terms that have used the
paraxial approximation.

If the tube lens is weakly focusing ($f_o \ll f_t$) then we can find the
electric field in the detector plane by taking the Fourier transform of the
electric field in the back focal plane
\begin{align}
  \mb{\tilde{e}}_d(\rd{}; \ro{}, \so{}) \propto \int_{\mathbb{R}^2} d\rb{}\, \tilde{\mb{e}}_b(\rb{}; \ro{}, \so{})\me^{-i (kn_t/f_t) \rb{} \cdot \rd{}}. \label{eq:ed}
\end{align}
By isolating the phase term in Eq. \ref{eq:bfp} with
$\mb{\tilde{e}}_b(\rb{};\ro{}, \so{}) \equiv \me^{-i(kn_o/f_o)\rb{}\cdot\ro{}}
\tilde{\underline{\mb{e}}}_b(\rb{};\so{})$, and plugging into Eq. \ref{eq:ed} we find
that 
\begin{align}
  \mb{\tilde{e}}_d(\rd{} - M\ro{}, \so{}) \propto \int_{\mathbb{R}^2} d\rb{}\, \tilde{\underline{\mb{e}}}_b(\rb{}; \so{})\me^{-i (kn_t/f_t)\, \rb{} \cdot [\,\rd{} - M\ro{}]}.
\end{align}
where $M = -\frac{n_o}{n_t}\frac{f_t}{f_o}$ is the transverse magnification.  By
writing the electric field in the detector plane in terms of $\rd{} - M\ro{}$,
we have established that the electric field in the detector plane is
\textit{transverse shift-invariant}---a transverse shift of the object creates a
magnified transverse shift of the image. We define coordinates on the detector
centered on the image of the object
$\rd{}' \equiv \rd{} - M\ro{} = r_d'\cos\phi_d'\mh{x} + r_d'\sin\phi_d'\mh{y}$,
then we follow Novotny \cite{nov2006} by writing the integrals in polar
coordinates, evaluating the azimuthal integrals, and writing the result
concisely in terms of three radial integrals
\begin{align}
  \tilde{\mb{e}}_d(\rd{}', \so{}) &\propto
  \begin{bmatrix}
    A(r_d') + C(r_d')\cos(2\phi_d') & C(r_d')\sin(2\phi_d') & 2iB(r_d')\cos\phi_d'\\
    C(r_d')\sin(2\phi_d') & A(r_d') - C(r_d')\cos(2\phi_d') & 2iB(r_d')\sin\phi_d'\\
    0&0&0\\
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix}\label{eq:elec}
\end{align}
where $A(r_d'), B(r_d')$, and $C(r_d')$ are
\begin{align}
  A(r_d') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin\theta(1+\cos\theta)J_0(k_b r_d'\sin\theta f_o/f_t),\\%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\},\\
  B(r_d') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin^2\theta J_1(k_b r_d'\sin\theta f_o/f_t),\\%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\},\\
  C(r_d') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin\theta(1-\cos\theta)J_2(k_b r_d'\sin\theta f_o/f_t).%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\}.
\end{align}
We can identify Eq. \ref{eq:elec} as the vector-valued \textit{coherent spread
  function} (CSF) of the microscope. The scalar-valued CSF---sometimes called
the amplitude transfer function---is only applicable to cases where
electromagnetic optics plays no role in the microscope---not true
for samples that contain dipole emitters.

To build our intuition about the CSF, we will rewrite the matrix multiplication
in Eq. \ref{eq:elec} in terms of the spherical harmonics. Notice that the $l=1$
spherical harmonics are the Cartesian components of the unit dipole axis (see
Eq. \ref{eq:harmonics}), so the CSF becomes
\begin{align}
  \tilde{\mb{e}}_d(\rd{}', \so{}) &\propto
  \begin{bmatrix}
    [A(r_d') + C(r_d')\cos(2\phi_d')]y_1^{1}(\so{}) + C(r_d')\sin(2\phi_d')y_1^{-1}(\so{}) + 2iB(r_d')\cos\phi_d'y_1^{0}(\so{})\\
    C(r_d')\sin(2\phi_d')y_1^{1}(\so{}) + [A(r_d') - C(r_d')\cos(2\phi_d')]y_1^{-1}(\so{}) + 2iB(r_d')\sin\phi_d'y_1^{0}(\so{})\\
    0\\
  \end{bmatrix}.
\end{align}
By applying the paraxial approximation ($\sin\theta\approx\theta$ and
$\cos\theta\approx 1$), the integrals $A(r_d'), B(r_d')$ and $C(r_d')$ can be
evaluated in terms of Bessel functions. We can evaluate $A(r_d')$ and $B(r_d')$ with the
help of $\int_0^zdx\ xJ_0(ax) = zJ_1(az)/a$ and
$\int_0^zdx\ x^2J_1(ax) = z^2J_2(az)/a$, respectively, and $C(r_d') = 0$ because
$J_2(x)$ is zero to first order. In this case, the CSF simplifies to
\begin{align}
  \tilde{\mb{e}}^{(p)}_d(\rd{}', \so{}) &\propto
  \begin{bmatrix}
    A^{(p)}(r_d')y_1^{1}(\so{}) + 2iB^{(p)}(r_d')\cos\phi_d'y_1^{0}(\so{})\\
    A^{(p)}(r_d')y_1^{-1}(\so{}) + 2iB^{(p)}(r_d')\sin\phi_d'y_1^{0}(\so{})\\
    0\\
  \end{bmatrix}\label{eq:paracsf}
\end{align}
where the integrals evaluate to
\begin{align}
  {A^{(p)}}(r_d') = \frac{J_1(2\pi ar_d')}{\pi ar_d'}, 
  &\hspace{2em}
    {B^{(p)}}(r_d') = \frac{\text{NA}}{n_o}\left[\frac{J_2(2\pi ar_d')}{\pi ar_d'}\right],  \label{eq:abparadef}
  \intertext{and we have substituted}
  a \equiv \frac{\text{NA}}{M\lambda},&\hspace{2em}
  \text{NA} \equiv n_o\sin\theta_{\text{max}}.
\end{align}
Under the paraxial approximation the electric fields on the detector created by
a single dipole are composed of two parts, a parallel part and a radial
part. The coefficients of the $y_1^0$ spherical harmonic in Eq. \ref{eq:paracsf}
create the radial part of the field---the $z$ component of the dipole generates
a radial field on the detector with respect to the image point. The coefficients
of the $y_1^1$ and $y_1^{-1}$ spherical harmonics create a parallel field in the
detector plane parallel to the $x$ and $y$ components of the
dipole. As the NA increases and the paraxial approximation no longer applies, we
begin to see coupling between the $x(y)$ component of the dipole with the $y(x)$
transverse field.

We can find the intensity in the detector plane due to a single dipole
by taking the modulus squared of the CSF
\begin{align}
  h(\rd{}', \so{}) = \left|\tilde{\mb{e}}_d(\rd{}', \so{})\right|^2 \label{eq:kernel}.
\end{align}
For convenience we keep the CSF written in terms of the spherical harmonics and
use a table of spherical harmonic products (see Appendix \ref{realspherical}) to
calculate the intensity in the detector plane as
\begin{equation}
  \begin{split}
  h(\rd{}', \so{}) \propto &\left(A^2(r_d') + 2B^2(r_d') + C^2(r_d')\right)y_0^0(\so{}) -\frac{2\sqrt{15}}{5}A(r_d')C(r_d')\sin(2\phi_d')y_2^{-2}(\so{})\\ &+ \frac{1}{\sqrt{5}}\left(-A^2(r_d') + 4B^2(r_d') - C^2(r_d')\right)y_2^{0}(\so{}) +\frac{2\sqrt{15}}{5}A(r_d')C(r_d')\cos(2\phi_d')y_2^{2}(\so{}). \label{eq:kerna}
\end{split}
\end{equation}
We can identify Eq. \ref{eq:kerna} as the spatio-angular \textit{point spread
  function} (PSF) of the microscope. Writing the spatio-angular PSF in terms of
spherical harmonic functions has two advantages. First, it allows us to express
the spatio-angular PSF very concisely. Instead of considering the point spread
function for every possible dipole orientation, we only need to consider four
spatio-angular PSFs---one for each spherical harmonic. Second, the spherical
harmonic functions form an orthonormal basis for functions on the sphere---a
convenient fact that we will use later.

@@

The spatio-angular PSF under the paraxial approximation is given by
\begin{align}
      h^{(p)}(\rd{}', \so{}) \propto \left({A^{(p)}}^2(r_d') + 2{B^{(p)}}^2(r_d')\right)y_0^0(\so{}) + \frac{1}{\sqrt{5}}\left(-{A^{(p)}}^2(r_d') + 4{B^{(p)}}^2(r_d')\right)y_2^{0}(\so{})\label{eq:para}.
\end{align}
First, consider the coefficient on the $y_0^0$ spherical harmonic in Eq.
\ref{eq:para}. This coefficient is the point spread function for an angularly
uniform distribution of fluorophores. The first term of the coefficient
${A^{(p)}}^2(r_d')$ is the familiar Airy disk that arises from the contribution
of dipoles oriented in the transverse plane, while the second term
${B^{(p)}}^2(r_d')$ is a smaller factor that arises from dipoles oriented
outside of the transverse plane. This leads to an interesting conclusion---a
uniform distribution of dipoles has a point spread function that is slightly
wider than an Airy disk even in the paraxial approximation. The Airy disk is
usually derived using paraxial scalar optics while here we have used paraxial
electromagnetic optics. Therefore, we can consider the second term to be an
electromagnetic correction to the Airy disk. We will quantify this difference in
the next section.

@@

Finally, consider all of the spherical harmonics that have a zero
coefficient. These spherical harmonics span the angular null space of our
measurement system---fluorophore distributions that lie in the null space do not
affect the measured intensities. Under the paraxial approximation all of the
non-zero coefficients are rotationally symmetric ($m=0$) spherical harmonics
which means that the transverse orientation of the dipoles does not affect the
PSF. In the high NA case this is no longer true---two $m=2$ spherical harmonics
have non-zero coefficients and the transverse orientation of dipoles does affect
the PSF.

@@

In Appendix \ref{paraxialctf} we calculate the paraxial CTF for a single-view fluorescence microscope as 
\begin{align}
  {\mb{E}_l^m}^{(p)}(\bs{\nu}) =
\begin{bmatrix}
  \delta(l-1, m-1) + \frac{2}{a}\frac{\text{NA}}{n_o}\nu\cos\phi_\nu\delta(l-1, m)\\
  \delta(l-1, m+1) + \frac{2}{a}\frac{\text{NA}}{n_o}\nu\sin\phi_\nu\delta(l-1, m)\\
  0\\
  \end{bmatrix}\Pi\left(\frac{\nu}{a}\right).\label{eq:ctfinal}
\end{align}
The CTF shows that single-view coherent microscopes have a spatial band limit at
$\nu = a$ and an $l=1$ angular pass band. It is useful to compare the paraxial
CTF in Eq. \ref{eq:ctfinal} to the paraxial electric field in the back focal
plane in Eq. \ref{eq:bfppara}. The CTF is just a rescaled version of the
electric field in the back focal plane, and this leads to an extremely valuable
interpretation. The CTF of a single-view fluorescence microscope is constrained
to the three members of the $l=1$ band the value of the CTF for each $m$ is a
rescaled version of the electric field due to a single dipole oriented along one
of the Cartesian axes.

@@

The spatial band limit can be increased by increasing the NA of the microscope,
so what parameter sets the angular band limit? The angular band limit in the
current microscope is set by the dipole emission process---only $l=1$ terms show
up in the CTF for dipole emitters. Adding polarizing filters to the microscope
will not extend the band limit of the microscope because filters only block
electric fields. However, polarized illumination can be used to extend the band
limit. The excitation process is completely incoherent with the emission
process, so the spatio-angular transfer functions of the two processes will
multiply. We can use the multiplication rules in Appendix \ref{realspherical} to
find the band limit of the new microscope that uses selective excitation.

>>>>

%%%%%%%%%%%%
% Document %
%%%%%%%%%%%%
\begin{document}
\title{\vspace{-2.5em} Spatio-angular transfer functions for fluorescence microscopes\vspace{-1em}}  \author{Talon Chandler, Min Guo, Hari Shroff, Rudolf Oldenbourg, Patrick La Rivi\`ere}
\date{\vspace{-1em}\today\vspace{-1em}}
\maketitle
\begin{abstract}
  We investigate how the orientation and position of fluorescent dipole emitters
  affects microscopic imaging using electromagnetic optics theory. Starting with
  the thoroughly studied spatio-angular point spread function, we introduce the
  spatio-angular coherent spread function, coherent transfer function, and
  optical transfer function as electromagnetic extensions of well-known
  functions in scalar optics theory. We use these concepts to show that
  fluorescence microscopes have a spatio-angular band limit. 
\end{abstract}
\section{Introduction}

@@

Figure \ref{fig:frames_a} shows a schematic of the fluorescence microscope that
we are considering with a summary of our notation. We start by following Backer
and Moerner \cite{backer2014} to find the electric field at position $\rb{}$ in
the back focal plane due to a single dipole emitter at position $\ro{}$ oriented
along $\so{}$ as
\begin{align}
  \mb{\tilde{e}}_b(\rb{};\ro{}, \so{}) \propto \me^{-i(kn_o/f_o)\rb{}\cdot\ro{}}\sqrt{\frac{1}{\rho_b}}
  \begin{bmatrix}
    \sin^2\phi_b + \rho_b\cos^2\phi_b&\sin\phi_b\cos\phi_b(\rho_b - 1)&-\frac{r_b}{f_o}\cos\phi_b\\
    \sin\phi_b\cos\phi_b(\rho_b - 1)&\cos^2\phi_b + \rho_b\sin^2\phi_b&-\frac{r_b}{f_o}\sin\phi_b\\
    0&0&0
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix}
\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)\label{eq:bfp},
\end{align}
where we define $\rho_b \equiv \sqrt{1 - \left(\frac{r_b}{f_o}\right)^2}$, and
$\Pi(x)$ is a boxcar function that returns 1 when $|x| < 1$ and 0 otherwise. We
can understand this expression term by term---the exponential term accounts for
the phase as dipole emitter is moved in object space, the square root term
conserves power before and after the objective lens, the matrix models the
dipole emission process and electric field rotation caused by the objective
lens, the vector is the dipole orientation unit vector, and
$\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right)$ accounts for the numerical
aperture of the lens with $r_b^{\text{max}} = \frac{f_0}{n_0}\text{NA}$. We use
a tilde to mark single dipole response functions---we will consider the response
due to a field of dipoles in the next section.

Although we will not use the result immediately, we can write the electric field
in the back focal plane under the paraxial approximation. We expand Eq. \ref{eq:bfp}
in a Taylor series about $r_b=0$ and drop all second-order and higher terms to
find
\begin{align}
  \mb{\tilde{e}}^{(p)}_b(\rb{};\ro{}, \so{}) \propto \me^{-i(kn_o/f_o)\rb{}\cdot\ro{}}
  \begin{bmatrix}
    1 & 0 &-\frac{r_b}{f_o}\cos\phi_b\\
    0&1&-\frac{r_b}{f_o}\sin\phi_b\\
    0&0&0
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix}
\Pi\left(\frac{r_b}{r_b^{\text{max}}}\right).\label{eq:bfppara}
\end{align}
We will continue to use a superscript $(p)$ to mark terms that have used the
paraxial approximation.

If the tube lens is weakly focusing ($f_o \ll f_t$) then we can find the
electric field in the detector plane by taking the Fourier transform of the
electric field in the back focal plane
\begin{align}
  \mb{\tilde{e}}_d(\rd{}; \ro{}, \so{}) \propto \int_{\mathbb{R}^2} d\rb{}\, \tilde{\mb{e}}_b(\rb{}; \ro{}, \so{})\me^{-i (kn_t/f_t) \rb{} \cdot \rd{}}. \label{eq:ed}
\end{align}
By isolating the phase term in Eq. \ref{eq:bfp} with
$\mb{\tilde{e}}_b(\rb{};\ro{}, \so{}) \equiv \me^{-i(kn_o/f_o)\rb{}\cdot\ro{}}
\tilde{\underline{\mb{e}}}_b(\rb{};\so{})$, and plugging into Eq. \ref{eq:ed} we find
that 
\begin{align}
  \mb{\tilde{e}}_d(\rd{} - M\ro{}, \so{}) \propto \int_{\mathbb{R}^2} d\rb{}\, \tilde{\underline{\mb{e}}}_b(\rb{}; \so{})\me^{-i (kn_t/f_t)\, \rb{} \cdot [\,\rd{} - M\ro{}]}.
\end{align}
where $M = -\frac{n_o}{n_t}\frac{f_t}{f_o}$ is the transverse magnification.  By
writing the electric field in the detector plane in terms of $\rd{} - M\ro{}$,
we have established that the electric field in the detector plane is
\textit{transverse shift-invariant}---a transverse shift of the object creates a
magnified transverse shift of the image. We define coordinates on the detector
centered on the image of the object
$\rd{}' \equiv \rd{} - M\ro{} = r_d'\cos\phi_d'\mh{x} + r_d'\sin\phi_d'\mh{y}$,
then we follow Novotny \cite{nov2006} by writing the integrals in polar
coordinates, evaluating the azimuthal integrals, and writing the result
concisely in terms of three radial integrals
\begin{align}
  \tilde{\mb{e}}_d(\rd{}', \so{}) &\propto
  \begin{bmatrix}
    a(r_d') + c(r_d')\cos(2\phi_d') & c(r_d')\sin(2\phi_d') & 2ib(r_d')\cos\phi_d'\\
    c(r_d')\sin(2\phi_d') & a(r_d') - c(r_d')\cos(2\phi_d') & 2ib(r_d')\sin\phi_d'\\
    0&0&0\\
  \end{bmatrix}
  \begin{bmatrix}
    \cos\varphi_o\sin\vartheta_o\\
    \sin\varphi_o\sin\vartheta_o\\
    \cos\vartheta_o
  \end{bmatrix},\label{eq:elec}
\end{align}
where $a(r_d'), b(r_d')$, and $c(r_d')$ are
\begin{align}
  a(r_d') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin\theta(1+\cos\theta)J_0(k_b r_d'\sin\theta f_o/f_t),\\%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\},\\
  b(r_d') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin^2\theta J_1(k_b r_d'\sin\theta f_o/f_t),\\%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\},\\
  c(r_d') &= \int_0^{\theta_{\text{max}}}d\theta\ \sqrt{\cos\theta}\sin\theta(1-\cos\theta)J_2(k_b r_d'\sin\theta f_o/f_t).%\ \text{exp}\left\{ik_bz_o[1-(1/2)(f_o/f_t)^2\sin^2\theta]\right\}.
\end{align}
We can identify Eq. \ref{eq:elec} as the vector-valued \textit{coherent spread
  function} (CSF) of the microscope. The more familiar scalar-valued
CSF---sometimes called the amplitude transfer function---is only applicable to
cases where electromagnetic optics plays no role in the microscope---not true
for samples that contain dipole emitters.

To build our intuition about the CSF, we will rewrite the matrix multiplication
in Eq. \ref{eq:elec} in terms of the spherical harmonics. Notice that the $l=1$
spherical harmonics are the Cartesian components of the unit dipole axis (see
Eq. \ref{eq:harmonics}), so the CSF becomes
\begin{align}
  \tilde{\mb{e}}_d(\rd{}', \so{}) &\propto
  \begin{bmatrix}
    [a(r_d') + c(r_d')\cos(2\phi_d')]y_1^{1}(\so{}) + c(r_d')\sin(2\phi_d')y_1^{-1}(\so{}) + 2ib(r_d')\cos\phi_d'y_1^{0}(\so{})\\
    c(r_d')\sin(2\phi_d')y_1^{1}(\so{}) + [a(r_d') - c(r_d')\cos(2\phi_d')]y_1^{-1}(\so{}) + 2ib(r_d')\sin\phi_d'y_1^{0}(\so{})\\
    0\\
  \end{bmatrix}.
\end{align}
By applying the paraxial approximation ($\sin\theta\approx\theta$ and
$\cos\theta\approx 1$), the integrals $a(r_d'), b(r_d')$ and $c(r_d')$ can be
evaluated in terms of Bessel functions. We can evaluate $a(r_d')$ and $b(r_d')$ with the
help of $\int_0^zdx\ xJ_0(ax) = zJ_1(az)/a$ and
$\int_0^zdx\ x^2J_1(ax) = z^2J_2(az)/a$, respectively, and $c(r_d') = 0$ because
$J_2(x)$ is zero to first order. In this case, the CSF simplifies to
\begin{align}
  \tilde{\mb{e}}^{(p)}_d(\rd{}', \so{}) &\propto
  \begin{bmatrix}
    a^{(p)}(r_d')y_1^{1}(\so{}) + 2ib^{(p)}(r_d')\cos\phi_d'y_1^{0}(\so{})\\
    a^{(p)}(r_d')y_1^{-1}(\so{}) + 2ib^{(p)}(r_d')\sin\phi_d'y_1^{0}(\so{})\\
    0\\
  \end{bmatrix}\label{eq:paracsf},
\end{align}
where the integrals evaluate to
\begin{align}
  {a^{(p)}}(r_d') = \frac{J_1(2\pi \nu_or_d')}{\pi \nu_or_d'}, 
  &\hspace{2em}
    {b^{(p)}}(r_d') = \frac{\text{NA}}{n_o}\left[\frac{J_2(2\pi \nu_or_d')}{\pi \nu_or_d'}\right],  \label{eq:abparadef}
  \intertext{and we have substituted}
  \nu_o \equiv \frac{\text{NA}}{M\lambda},&\hspace{2em}
  \text{NA} \equiv n_o\sin\theta_{\text{max}}.
\end{align}
Under the paraxial approximation the electric fields on the detector created by
a single dipole are composed of two parts, a parallel part and a radial
part. The coefficients of the $y_1^0$ spherical harmonic in Eq. \ref{eq:paracsf}
create the radial part of the field---the $z$ component of the dipole generates
a radial field on the detector with respect to the image point. The coefficients
of the $y_1^1$ and $y_1^{-1}$ spherical harmonics create a parallel field in the
detector plane parallel to the $x$ and $y$ components of the
dipole. As the NA increases and the paraxial approximation no longer applies, we
begin to see coupling between the $x$ and $y$ components of the dipole with the $y$ and $x$ transverse fields, respectively.

We can find the intensity in the detector plane due to a single dipole
by taking the squared modulus of the CSF
\begin{align}
  h(\rd{}', \so{}) = \left|\tilde{\mb{e}}_d(\rd{}', \so{})\right|^2 \label{eq:kernel}.
\end{align}
For convenience we keep the CSF written in terms of the spherical harmonics and
use a table of spherical harmonic products (see Appendix \ref{realspherical}) to
calculate the intensity in the detector plane as
\begin{equation}
  \begin{split}
  h(\rd{}', \so{}) \propto &\left(a^2(r_d') + 2b^2(r_d') + c^2(r_d')\right)y_0^0(\so{}) -\frac{2\sqrt{15}}{5}a(r_d')c(r_d')\sin(2\phi_d')y_2^{-2}(\so{})\\ &+ \frac{1}{\sqrt{5}}\left(-a^2(r_d') + 4b^2(r_d') - c^2(r_d')\right)y_2^{0}(\so{}) +\frac{2\sqrt{15}}{5}a(r_d')c(r_d')\cos(2\phi_d')y_2^{2}(\so{}). \label{eq:kerna}
\end{split}
\end{equation}
We can identify Eq. \ref{eq:kerna} as the spatio-angular \textit{point spread
  function} (PSF) of the microscope. Writing the spatio-angular PSF in terms of
spherical harmonic functions has two advantages. First, it allows us to express
the spatio-angular PSF very concisely. Instead of considering the point spread
function for every possible dipole orientation, we only need to consider four
spatio-angular PSFs---one for each spherical harmonic. Second, the spherical
harmonic functions form an orthonormal basis for functions on the sphere---a
convenient fact that we will use later.

@@

The spatio-angular PSF under the paraxial approximation is given by
\begin{align}
      h^{(p)}(\rd{}', \so{}) \propto \left({a^{(p)}}^2(r_d') + 2{b^{(p)}}^2(r_d')\right)y_0^0(\so{}) + \frac{1}{\sqrt{5}}\left(-{a^{(p)}}^2(r_d') + {4b^{(p)}}^2(r_d')\right)y_2^{0}(\so{})\label{eq:para}.
\end{align}
We rewrite the PSF in the following form to make plotting and interpretation easier
\begin{align}
  h^{(p)}(\rd{}', \so{}) &= {h_0^0}^{(p)}(\rd{}')y_0^0(\so{}) + {h_0^2}^{(p)}(\rd{}')y_0^2(\so{}),\\
  {h_0^0}^{(p)}(\rd{}') &\equiv {a^{(p)}}^2(r_d') + 2{b^{(p)}}^2(r_d'),\\
  {h_0^2}^{(p)}(\rd{}') &\equiv \frac{1}{\sqrt{5}}\left[- {a^{(p)}}^2(r_d') + 4{b^{(p)}}^2(r_d')\right].
\end{align}

First, consider the coefficient on the $y_0^0$ spherical harmonic in Eq.
\ref{eq:para}. This coefficient is the point spread function for an angularly
uniform distribution of fluorophores. The first term of the coefficient
${A^{(p)}}^2(r_d')$ is the familiar Airy disk that arises from the contribution
of dipoles oriented in the transverse plane, while the second term
${B^{(p)}}^2(r_d')$ is a smaller factor that arises from dipoles oriented
outside of the transverse plane. This leads to an interesting conclusion---a
uniform distribution of dipoles has a point spread function that is slightly
wider than an Airy disk even in the paraxial approximation. The Airy disk is
usually derived using paraxial scalar optics while here we have used paraxial
electromagnetic optics. Therefore, we can consider the second term to be an
electromagnetic correction to the Airy disk. We will quantify this difference in
the next section.

@@

Finally, consider all of the spherical harmonics that have a zero
coefficient. These spherical harmonics span the angular null space of our
measurement system---fluorophore distributions that lie in the null space do not
affect the measured intensities. Under the paraxial approximation all of the
non-zero coefficients are rotationally symmetric ($m=0$) spherical harmonics,
which means that the transverse orientation of the dipoles does not affect the
PSF. In the high NA case this is no longer true---two $m=2$ spherical harmonics
have non-zero coefficients and the transverse orientation of dipoles does affect
the PSF.

@@

In Appendix \ref{paraxialctf} we calculate the paraxial CTF for a single-view fluorescence microscope as 
\begin{align}
  {\mb{E}_l^m}^{(p)}(\bs{\nu}) =
\begin{bmatrix}
  \delta(l-1, m-1) + \frac{2}{\nu_o}\frac{\text{NA}}{n_o}\nu\cos\phi_\nu\delta(l-1, m)\\
  \delta(l-1, m+1) + \frac{2}{\nu_o}\frac{\text{NA}}{n_o}\nu\sin\phi_\nu\delta(l-1, m)\\
  0\\
  \end{bmatrix}\Pi\left(\frac{\nu}{\nu_o}\right).\label{eq:ctfinal}
\end{align}
The CTF shows that single-view coherent microscopes have a spatial band limit at
$\nu = \nu_o$ and an $l=1$ angular pass band. It is useful to compare the paraxial
CTF in Eq. \ref{eq:ctfinal} to the paraxial electric field in the back focal
plane in Eq. \ref{eq:bfppara}. The CTF is just a rescaled version of the
electric field in the back focal plane, and this leads to an extremely valuable
interpretation. The CTF of a single-view fluorescence microscope is constrained
to the three members of the $l=1$ band. The value of the CTF for each $m$ is a
rescaled version of the electric field due to a single dipole oriented along one
of the Cartesian axes.

@@

The spatial band limit can be increased by increasing the NA of the microscope,
so what parameter sets the angular band limit? The angular band limit in the
current microscope is set by the dipole emission process---only $l=1$ terms show
up in the CTF for dipole emitters. Adding polarizing filters to the microscope
will not extend the band limit of the microscope because filters only block
electric fields. However, polarized illumination can be used to extend the band
limit. The excitation process is completely incoherent with the emission
process, so the spatio-angular transfer functions of the two processes will
multiply. We can use the multiplication rules in Appendix \ref{realspherical} to
find the band limit of a new microscope that uses selective excitation.

Next we find the intensity on the detector plane $g(\rd{})$ by taking the
modulus squared of the electric field. Using Eq. \ref{eq:eforward} we find that
\begin{align}
  g(\rd{}) = |\mb{e}_d(\rd{})|^2 = \left|\int_{\mathbb{S}^2}d\so{}\int_{\mathbb{R}^2}d\ro{} \ \tilde{\mb{e}}_d(\rd{} - M\mb{r}_o, \so{})\bs{\mu}(\ro{}, \so{})\right|^2. 
\end{align}
If the dipoles emit incoherently---a fair assumption for fluorescent emitters
that are not within homo-FRET distance---then we can take the squared modulus of the
object and CSF independently, which gives
\begin{align}
  g(\rd{}) = \int_{\mathbb{S}^2}d\so{}\int_{\mathbb{R}^2}d\ro{} \ \left|\tilde{\mb{e}}_d(\rd{} - M\mb{r}_o, \so{})\right|^2\left|\bs{\mu}(\ro{}, \so{})\right|^2. 
\end{align}
We recognize that $\left|\tilde{\mb{e}}_d(\rd{} - M\mb{r}_o, \so{})\right|^2$ is
the PSF of the microscope $h(\rd{} - M\mb{r}_o, \so{})$, and
we introduce a new function
$f(\ro{}, \so{}) \equiv \left|\bs{\mu}(\ro{}, \so{})\right|^2$ to represent the object
\begin{align}
  g(\rd{}) = \int_{\mathbb{S}^2}d\so{}\int_{\mathbb{R}^2}d\ro{} \ h(\rd{} - M\mb{r}_o, \so{})f(\ro{}, \so{}).\label{eq:forwardint}
\end{align}
We can interpret $f(\ro{}, \so{})$ as the spatio-angular density of dipoles---it
is a function that is proportional to the number of fluorophores at position
$\ro{}$ oriented in direction $\so{}$ per unit volume per unit solid angle. Notice
that $h$ and $f$ are scalar functions---we removed all phase information when we took
the modulus squared.
